# Week 6: Hypothesis Testing^[Adapted nearly verbatim from Chapter 11 in Navarro, D. ["Learning Statistics with R."](https://learningstatisticswithr.com) R code from Emily Kothe's Bookdown adaptation of [Learning statistics with R](https://github.com/ekothe/rbook)]

>*The process of induction is the process of assuming the simplest law that can be made to harmonize with our experience. This process, however, has no logical foundation but only a psychological one. It is clear that there are no grounds for believing that the simplest course of events will really happen. It is an hypothesis that the sun will rise tomorrow: and this means that we do not know whether it will rise.*
>
>-- Ludwig Wittgenstein^[The quote comes from Wittgenstein's (1922) text, *Tractatus Logico-Philosphicus*.]

In the last chapter, we discussed estimation and sampling theory: "big ideas" in inferential statistics. It's now time to turn our attention to **hypothesis testing**. In its most abstract form, hypothesis testing really a very simple idea: the researcher has some theory about the world, and wants to determine whether or not the data actually support that theory. However, the details are messy, and most people find the theory of hypothesis testing to be the most frustrating part of statistics. The structure of the chapter is as follows. Firstly, I'll describe how hypothesis testing works, in a fair amount of detail, using a simple running example to show you how a hypothesis test is "built". I'll try to avoid being too dogmatic while doing so, and focus instead on the underlying logic of the testing procedure.^[A technical note. The description below differs subtly from the standard description given in a lot of introductory texts. The orthodox theory of null hypothesis testing emerged from the work of Sir Ronald Fisher and Jerzy Neyman in the early 20th century; but Fisher and Neyman actually had very different views about how it should work. The standard treatment of hypothesis testing that most texts use is a hybrid of the two approaches. The treatment here is a little more Neyman-style than the orthodox view, especially as regards the meaning of the $p$ value.] Afterwards, I'll spend a bit of time talking about the various dogmas, rules and heresies that surround the theory of hypothesis testing. 

## A menagerie of hypotheses{#hypotheses}
Eventually we all succumb to madness. For me, that day will arrive once I'm finally promoted to full professor. Safely ensconced in my ivory tower, happily protected by tenure, I will finally be able to take leave of my senses (so to speak), and indulge in that most thoroughly unproductive line of psychological research: the search for extrasensory perception (ESP).^[My apologies to anyone who actually believes in this stuff, but on my reading of the literature on ESP, it's just not reasonable to think this is real. To be fair, though, some of the studies are rigorously designed; so it's actually an interesting area for thinking about psychological research design. And of course it's a free country, so you can spend your own time and effort proving me wrong if you like, but I wouldn't think that's a terribly practical use of your intellect.] 

Let's suppose that this glorious day has come. My first study is a simple one, in which I seek to test whether clairvoyance exists. Each participant sits down at a table, and is shown a card by an experimenter. The card is black on one side and white on the other. The experimenter takes the card away, and places it on a table in an adjacent room. The card is placed black side up or white side up completely at random, with the randomisation occurring only after the experimenter has left the room with the participant. A second experimenter comes in and asks the participant which side of the card is now facing upwards. It's purely a one-shot experiment. Each person sees only one card, and gives only one answer; and at no stage is the participant actually in contact with someone who knows the right answer. My data set, therefore, is very simple. I have asked the question of $N$ people, and some number $X$ of these people have given the correct response. To make things concrete, let's suppose that I have tested $N = 100$ people, and $X = 62$ of these got the answer right... a surprisingly large number, sure, but is it large enough for me to feel safe in claiming I've found evidence for extrasensory perception? This is the situation where hypothesis testing comes in useful. However, before we talk about how to *test* hypotheses, we need to be clear about what we mean by hypotheses.

### Research hypotheses versus statistical hypotheses
The first distinction that you need to keep clear in your mind is between **research hypotheses** and **statistical hypotheses**. In my ESP study, my overall scientific goal is to demonstrate that clairvoyance exists. In this situation, I have a clear research goal: I am hoping to discover evidence for ESP. In other situations I might actually be a lot more neutral than that, so I might say that my research goal is to determine whether or not clairvoyance exists. Regardless of how I want to portray myself, the basic point that I'm trying to convey here is that a research hypothesis involves making a substantive, testable scientific claim... if you are a psychologist, then your research hypotheses are fundamentally *about* psychological constructs. Any of the following would count as **research hypotheses**:

- **Listening to music reduces your ability to pay attention to other things**. This is a claim about the causal relationship between two psychologically meaningful concepts (listening to music and paying attention to things), so it's a perfectly reasonable research hypothesis.
- **Intelligence is related to personality**. Like the last one, this is a relational claim about two psychological constructs (intelligence and personality), but the claim is weaker: correlation is not causation.
- **Intelligence is speed of information processing**. This hypothesis has a quite different character: it's not actually a relational claim at all. It's an ontological claim about the fundamental character of intelligence (and I'm pretty sure it's wrong). 

It's worth expanding on the last research hypothesis actually. It's usually easier to think about how to construct experiments to test research hypotheses of the form "does X affect Y?" than it is to address claims like "what is X?" And in practice, what usually happens is that you find ways of testing relational claims that follow from your ontological ones. For instance, if I believe that intelligence *is* speed of information processing in the brain, my experiments will often involve looking for relationships between measures of intelligence and measures of speed. As a consequence, most everyday research questions do tend to be relational in nature, but they're almost always motivated by deeper ontological questions about the state of nature. 

Notice that in practice, my research hypotheses could overlap a lot. My ultimate goal in the ESP experiment might be to test an ontological claim like "ESP exists", but I might operationally restrict myself to a narrower hypothesis like "Some people can `see' objects in a clairvoyant fashion". That said, there are some things that really don't count as proper research hypotheses in any meaningful sense: 

- **Love is a battlefield**. This is too vague to be testable. While it's okay for a research hypothesis to have a degree of vagueness to it, it has to be possible to operationalise your theoretical ideas. Maybe I'm just not creative enough to see it, but I can't see how this can be converted into any concrete research design. If that's true, then this isn't a scientific research hypothesis, it's a pop song. That doesn't mean it's not interesting -- a lot of deep questions that humans have fall into this category. Maybe one day science will be able to construct testable theories of love, or to test to see if God exists, and so on; but right now we can't, and I wouldn't bet on ever seeing a satisfying scientific approach to either. 
- **The first rule of tautology club is the first rule of tautology club**. This is not a substantive claim of any kind. It's true by definition. No conceivable state of nature could possibly be inconsistent with this claim. As such, we say that this is an unfalsifiable hypothesis, and as such it is outside the domain of science. Whatever else you do in science, your claims must have the possibility of being wrong. 
- **More people in my experiment will say "yes" than "no"**. This one fails as a research hypothesis because it's a claim about the data set, not about the psychology (unless of course your actual research question is whether people have some kind of "yes" bias!). As we'll see shortly, this hypothesis is starting to sound more like a statistical hypothesis than a research hypothesis. 


As you can see, research hypotheses can be somewhat messy at times; and ultimately they are *scientific* claims. **Statistical hypotheses** are neither of these two things. Statistical hypotheses must be mathematically precise, and they must correspond to specific claims about the characteristics of the data generating mechanism (i.e., the "population"). Even so, the intent is that statistical hypotheses bear a clear relationship to the substantive research hypotheses that you care about! For instance, in my ESP study my research hypothesis is that some people are able to see through walls or whatever. What I want to do is to "map" this onto a statement about how the data were generated. So let's think about what that statement would be. The quantity that I'm interested in within the experiment is $P(\mbox{correct})$, the true-but-unknown **probability** with which the participants in my experiment answer the question correctly. Let's use the Greek letter $\theta$ (theta) to refer to this probability. Here are four different statistical hypotheses:

- If ESP doesn't exist and if my experiment is well designed, then my participants are just guessing. So I should expect them to get it right half of the time and so my statistical hypothesis is that the true probability of choosing correctly is $\theta = 0.5$. 
- Alternatively, suppose ESP does exist and participants can see the card. If that's true, people will perform better than chance. The statistical hypothesis would be that $\theta > 0.5$. 
- A third possibility is that ESP does exist, but the colors are all reversed and people don't realise it (okay, that's wacky, but you never know...). If that's how it works then you'd expect people's performance to be *below* chance. This would correspond to a statistical hypothesis that $\theta < 0.5$. 
- Finally, suppose ESP exists, but I have no idea whether people are seeing the right color or the wrong one. In that case, the only claim I could make about the data would be that the probability of making the correct answer is *not* equal to 50. This corresponds to the statistical hypothesis that $\theta \neq 0.5$. 

All of these are legitimate examples of a statistical hypothesis because they are statements about a **population parameter** and are meaningfully related to my experiment.

What this discussion makes clear, I hope, is that when attempting to construct a statistical hypothesis test the researcher actually has two quite distinct hypotheses to consider. First, he or she has a research hypothesis (a claim about something), and this corresponds to a statistical hypothesis (a claim about the data generating population). In my ESP example, these might be:

```{r echo=FALSE}
knitr::kable(data.frame(stringsAsFactors=FALSE,
   `My research hypothesis` = c("ESP exists"),
   `My statistical hypothesis` = c("$\\theta \\neq 0.5$")), 
   col.names = c("My research hypothesis","My statistical hypothesis"))
```

The key thing to recognise is this: *a statistical hypothesis test is a test of the statistical hypothesis, not the research hypothesis*. If your study is badly designed, then the link between your research hypothesis and your statistical hypothesis is broken. To give a silly example, suppose that my ESP study was conducted in a situation where the participant can actually see the card reflected in a window; if that happens, I would be able to find very strong evidence that $\theta \neq 0.5$, but this would tell us nothing about whether "ESP exists". 

### Null hypotheses and alternative hypotheses
So far, so good. I have a research hypothesis that corresponds to what I want to **believe about the world**, and I can map it onto a statistical hypothesis that corresponds to what I want to **believe about how the data were generated**. It's at this point that things get somewhat counterintuitive for a lot of people. Because what I'm about to do is invent a new statistical hypothesis (the "null" hypothesis, $H_0$) that corresponds to the exact opposite of what I want to believe, and then focus exclusively on that, almost to the neglect of the thing I'm actually interested in (the "alternative" hypothesis: $H_1$). In our ESP example, the null hypothesis is that $\theta = 0.5$, since that's what we'd expect if ESP *didn't* exist. My hope, of course, is that ESP is totally real, and so the *alternative* to this null hypothesis is $\theta \neq 0.5$. In essence, what we're doing here is dividing up the possible values of $\theta$ into two groups: those values that I really hope aren't true (the null), and those values that I'd be happy with if they turn out to be right (the alternative). Having done so, the important thing to recognise is that the goal of a hypothesis test is *not* to show that the alternative hypothesis is (probably) true; the goal is to show that the null hypothesis is (probably) false. Most people find this pretty weird. 

The best way to think about it, in my experience, is to imagine that a hypothesis test is a criminal trial^[This analogy only works if you're from an adversarial legal system like UK/US/Australia. As I understand these things, the French inquisitorial system is quite different.]: *the trial of the null hypothesis*. The null hypothesis is the defendant, the researcher is the prosecutor, and the statistical test itself is the judge. Just like a criminal trial, there is a presumption of innocence: the null hypothesis is *deemed* to be true unless you, the researcher, can prove beyond a reasonable doubt that it is false. You are free to design your experiment however you like (within reason, obviously!), and your goal when doing so is to maximise the chance that the data will yield a conviction... for the crime of being false. The catch is that the statistical test sets the rules of the trial, and those rules are designed to protect the null hypothesis -- specifically to ensure that if the null hypothesis is actually true, the chances of a false conviction are guaranteed to be low. This is pretty important: after all, the null hypothesis doesn't get a lawyer. And given that the researcher is trying desperately to prove it to be false, *someone* has to protect it.  



## Two types of errors{#errortypes}
Before going into details about how a statistical test is constructed, it's useful to understand the philosophy behind it. I hinted at it when pointing out the similarity between a null hypothesis test and a criminal trial, but I should now be explicit. Ideally, we would like to construct our test so that we never make any errors. Unfortunately, since the world is messy, this is never possible. Sometimes you're just really unlucky: for instance, suppose you flip a coin 10 times in a row and it comes up heads all 10 times. That feels like very strong evidence that the coin is biased, but of course there's a 1 in 1024 chance that this would happen even if the coin was totally fair ($\frac{1}{0.5^{10}}$ or try it in R: `dbinom(10,10,0.5)`). In other words, in real life we *always* have to accept that there's a chance that we did the wrong thing. As a consequence, the goal behind statistical hypothesis testing is not to *eliminate* errors, but to *minimise* them.

At this point, we need to be a bit more precise about what we mean by "errors". Firstly,  let's state the obvious: it is either the case that the null hypothesis is true, or it is false; and our test will either reject the null hypothesis or retain it.^[An aside regarding the language you use to talk about hypothesis testing. Firstly, one thing you really want to avoid is the word "prove": a statistical test really doesn't *prove* that a hypothesis is true or false. Proof implies certainty, and as the saying goes, statistics means never having to say you're certain. On that point almost everyone would agree. However, beyond that there's a fair amount of confusion. Some people argue that you're only allowed to make statements like "rejected the null", "failed to reject the null", or possibly "retained the null". According to this line of thinking, you can't say things like "accept the alternative" or "accept the null". Personally I think this is too strong: in my opinion, this conflates null hypothesis testing with Karl Popper's falsificationist view of the scientific process. While there are similarities between falsificationism and null hypothesis testing, they aren't equivalent. However, while I personally think it's fine to talk about accepting a hypothesis (on the proviso that "acceptance" doesn't actually mean that it's necessarily true, especially in the case of the null hypothesis), many people will disagree. And more to the point, you should be aware that this particular weirdness exists, so that you're not caught unaware by it when writing up your own results.] So, as the table below illustrates, after we run the test and make our choice, one of four things might have happened:

```{r echo=FALSE}
knitr::kable(data.frame(stringsAsFactors=FALSE,
NANA = c("$H_0$ is true", "$H_0$ is false"),
`retain.$H_0$` = c("correct decision", "error (type II)"),
`retain.$H_0$` = c("error (type I)", "correct decision")
), col.names = c("", "retain $H_0$", "reject $H_0$"))

```



As a consequence there are actually *two* different types of error here. If we reject a null hypothesis that is actually true, then we have made a **type I error**. On the other hand, if we retain the null hypothesis when it is in fact false, then we have made a **type II error**. 

Remember how I said that statistical testing was kind of like a criminal trial? Well, I meant it. A criminal trial requires that you establish "beyond a reasonable doubt" that the defendant did it. All of the evidentiary rules are (in theory, at least) designed to ensure that there's (almost) no chance of wrongfully convicting an innocent defendant. The trial is designed to protect the rights of a defendant: as the English jurist William Blackstone famously said, it is "better that ten guilty persons escape than that one innocent suffer." In other words, a criminal trial doesn't treat the two types of error in the same way~... punishing the innocent is deemed to be much worse than letting the guilty go free. A statistical test is pretty much the same: the single most important design principle of the test is to *control* the probability of a type I error, to keep it below some fixed probability. This probability, which is denoted $\alpha$, is called the **significance level** of the test. And I'll say it again, because it is so central to the whole set-up... a hypothesis test is said to have significance level $\alpha$ if the type I error rate is no larger than $\alpha$. 

So, what about the type II error rate? Well, we'd also like to keep those under control too, and we denote this probability by $\beta$. However, it's much more common to refer to the  **power** of the test, which is the probability with which we reject a null hypothesis when it really is false, which is $1-\beta$. To help keep this straight, here's the same table again, but with the relevant numbers added:

```{r echo=FALSE}
knitr::kable(data.frame(stringsAsFactors=FALSE,
NANA = c("$H_0$ is true", "$H_0$ is false"),
retain = c("$1-\\alpha$ (probability of correct retention)",
                   "$\\beta$ (type II error rate)"),
reject = c("$\\alpha$ (type I error rate)",
                   "$1-\\beta$  (power of the test)")), col.names=c("", "retain $H_0$", "reject $H_0$")
)

```



A "powerful" hypothesis test is one that has a small value of $\beta$, while still keeping $\alpha$ fixed at some (small) desired level. By convention, scientists mostly make use of three different $\alpha$ levels: $.05$, $.01$ and $.001$. Notice the asymmetry here... the tests are designed to *ensure* that the $\alpha$ level is kept small, but there's no corresponding guarantee regarding $\beta$. We'd certainly *like* the type II error rate to be small, and we try to design tests that keep it small, but this is very much secondary to the overwhelming need to control the type I error rate. As Blackstone might have said if he were a statistician, it is "better to retain 10 false null hypotheses than to reject a single true one". To be honest, I don't know that I agree with this philosophy -- there are situations where I think it makes sense, and situations where I think it doesn't -- but that's neither here nor there. It's how the tests are built.

## Test statistics and sampling distributions{#teststatistics} 
At this point we need to start talking specifics about how a hypothesis test is constructed. To that end, let's return to the ESP example. Let's ignore the actual data that we obtained, for the moment, and think about the structure of the experiment. Regardless of what the actual numbers are, the *form* of the data is that $X$ out of $N$ people correctly identified the colour of the hidden card. Moreover, let's suppose for the moment that the null hypothesis really is true: ESP doesn't exist, and the true probability that anyone picks the correct colour is exactly $\theta = 0.5$. What would we *expect* the data to look like? Well, obviously, we'd expect the proportion of people who make the correct response to be pretty close to 50\%. Or, to phrase this in more mathematical terms, we'd say that $X/N$ is approximately $0.5$. Of course, we wouldn't expect this fraction to be *exactly* 0.5: if, for example we tested $N=100$ people, and $X = 53$ of them got the question right, we'd probably be forced to concede that the data are quite consistent with the null hypothesis. On the other hand, if $X = 99$ of our participants got the question right, then we'd feel pretty confident that the null hypothesis is wrong. Similarly, if only $X=3$ people got the answer right, we'd be similarly confident that the null was wrong. Let's be a little more technical about this: we have a quantity $X$ that we can calculate by looking at our data; after looking at the value of $X$, we make a decision about whether to believe that the null hypothesis is correct, or to reject the null hypothesis in favour of the alternative. The name for this thing that we calculate to guide our choices is a **test statistic**. 

Having chosen a test statistic, the next step is to state precisely which values of the test statistic would cause to reject the null hypothesis, and which values would cause us to keep it. In order to do so, we need to determine what the **sampling distribution of the test statistic** would be if the null hypothesis were actually true (we talked about sampling distributions earlier in Section \@ref(sampdistmain)). Why do we need this? Because this distribution tells us exactly what values of $X$ our null hypothesis would lead us to expect. And therefore, we can use this distribution as a tool for assessing how closely the null hypothesis agrees with our data. 

```{r samplingdist, fig.cap="The sampling distribution for our test statistic $X$ when the null hypothesis is true. For our ESP scenario, this is a binomial distribution. Not surprisingly, since the null hypothesis says that the probability of a correct response is $\\theta = .5$, the sampling distribution says that the most likely value is 50 (our of 100) correct responses. Most of the probability mass lies between 40 and 60.", echo=FALSE}


nhstImg <- list()
emphCol <- rgb(0,0,1)
emphColLight <- rgb(.5,.5,1)
emphGrey <- grey(.5)

eps <- TRUE
colour <- TRUE
	width <- 8
	height <- 5.5
	
	# distribution
	x <- 0:100
	y <- dbinom(x,100,.5)
	
	# plot
	plot(x,y,type="h",lwd=3, col=ifelse(colour,emphCol,"black"),
		xlab="Number of Correct Responses (X)", ylab="Probability",
		main="Sampling Distribution for X if the Null is True",
		font.main=1, frame.plot=FALSE
	)
```

How do we actually determine the sampling distribution of the test statistic? For a lot of hypothesis tests this step is actually quite complicated, but fortunately for us, our ESP example provides us with one of the easiest cases. Our population parameter $\theta$ is just the overall probability that people respond correctly when asked the question, and our test statistic $X$ is the **count** of the number of people who did so, out of a sample size of $N$. We've seen a distribution like this before, in Section \@ref(binomdist): that's exactly what the binomial distribution describes! So, to use the notation and terminology that I introduced in that section, we would say that the null hypothesis predicts that $X$ is binomially distributed, which is written
$$
X \sim \mbox{Binomial}(\theta,N)
$$
Since the null hypothesis states that $\theta = 0.5$ and our experiment has $N=100$ people, we have the sampling distribution we need. This sampling distribution is plotted in Figure \@ref(fig:samplingdist). No surprises really: the null hypothesis says that $X=50$ is the most likely outcome, and it says that we're almost certain to see somewhere between 40 and 60 correct responses. 

## Making decisions{#decisionmaking}
Okay, we're very close to being finished. We've constructed a test statistic ($X$), and we chose this test statistic in such a way that we're pretty confident that if $X$ is close to $N/2$ then we should retain the null, and  if not we should reject it. The question that remains is this: exactly which values of the test statistic should we associate with the null hypothesis, and which exactly values go with the alternative hypothesis? In my ESP study, for example, I've observed a value of $X=62$. What decision should I make? Should I choose to believe the null hypothesis, or the alternative hypothesis?

### Critical regions and critical values
To answer this question, we need to introduce the concept of a **critical region** for the test statistic $X$. The critical region of the test corresponds to those values of $X$ that would lead us to reject null hypothesis (which is why the critical region is also sometimes called the rejection region). How do we find this critical region? Well, let's consider what we know: 

- $X$ should be very big or very small in order to reject the null hypothesis.
- If the null hypothesis is true, the sampling distribution of $X$ is Binomial$(0.5, N)$.
- If $\alpha =.05$, the critical region must cover 5\% of this sampling distribution. 

It's important to make sure you understand this last point: the critical region corresponds to those values of $X$ for which we would reject the null hypothesis, and the sampling distribution in question describes the probability that we would obtain a particular value of $X$ if the null hypothesis were actually true. Now, let's suppose that we chose a critical region that covers 20\% of the sampling distribution, and suppose that the null hypothesis is actually true. What would be the probability of incorrectly rejecting the null? The answer is of course 20\%. And therefore, we would have built a test that had an $\alpha$ level of $0.2$. If we want $\alpha = .05$, the critical region is only *allowed* to cover 5\% of the sampling distribution of our test statistic.

```{r crit2, fig.cap="The critical region associated with the hypothesis test for the ESP study, for a hypothesis test with a significance level of $\\alpha = .05$. The plot itself shows the sampling distribution of $X$ under the null hypothesis: the grey bars correspond to those values of $X$ for which we would retain the null hypothesis. The blue bars show the critical region: those values of $X$ for which we would reject the null. Because the alternative hypothesis is two sided (i.e., allows both $\\theta <.5$ and $\\theta >.5$), the critical region covers both tails of the distribution. To ensure an $\\alpha$ level of $.05$, we need to ensure that each of the two regions encompasses 2.5% of the sampling distribution.", echo=FALSE}
	# needed for printing
	width <- 8
	height <- 5.5
	
	setUpPlot <- function() {
		
		plot.new()
		plot.window(xlim=c(0,100),ylim=c(0,.08))
		axis(1)
		title(xlab="Number of Correct Responses (X)")
		
	}
	
	addDistPlot <- function(x,y,z) {
		
		# colour key
		col.key <- c(
			 grey(.9),
			 ifelse(colour,emphCol,"black")
		)
		
		# plot
		lines(x,y,col=col.key[as.numeric(z)+1],type="h",lwd=3)
		
	}
	
	addArrow <- function(x,h, text) {
		
		arrows(x0 = x[1], y0 = h, x1 = x[2], y1 = h, length = .1)
		lines(c(x[1],x[1]),c(h-.002,h+.002),'type'="l")
		
	}
	
	# distribution
	x <- 0:100
	y <- dbinom(x,100,.5)
	
	# plot 1
	setUpPlot()
	z <- x<=40 | x>=60
	addDistPlot(x,y,z)
	h <- .03
	addArrow(c(40,20),h)
	addArrow(c(60,80),h)	
	text(22,h+.013,"lower critical region")
	text(22,h+.007,"(2.5% of the distribution)")
	text(75,h+.013,"upper critical region")
	text(75,h+.007,"(2.5% of the distribution)")
	title(main="Critical Regions for a Two-Sided Test",font.main=1)
```

As it turns out, those three things uniquely solve the problem: our critical region consists of the most *extreme values*, known as the **tails** of the distribution. This is illustrated in Figure \@ref(fig:crit2). As it turns out, if we want $\alpha = .05$, then our critical regions correspond to $X \leq 40$ and $X \geq 60$.^[Strictly speaking, the test I just constructed has $\alpha = .057$, which is a bit too generous. However, if I'd chosen 39 and 61 to be the boundaries for the critical region, then the critical region only covers 3.5\% of the distribution. I figured that it makes more sense to use 40 and 60 as my critical values, and be willing to tolerate a 5.7\% type I error rate, since that's as close as I can get to a value of $\alpha = .05$.] That is, if the number of people saying "true" is between 41 and 59, then we should retain the null hypothesis. If the number is between 0 to 40 or between 60 to 100, then we should reject the null hypothesis. The numbers 40 and 60 are often referred to as the **critical values**, since they define the edges of the critical region.

At this point, our hypothesis test is essentially complete: (1) we choose an $\alpha$ level (e.g., $\alpha = .05$, (2) come up with some test statistic (e.g., $X$) that does a good job (in some meaningful sense) of comparing $H_0$ to $H_1$, (3) figure out the sampling distribution of the test statistic on the assumption that the null hypothesis is true (in this case, binomial) and then (4) calculate the critical region that produces an appropriate $\alpha$ level (0-40 and 60-100). All that we have to do now is calculate the value of the test statistic for the real data (e.g., $X = 62$) and then compare it to the critical values to make our decision. Since 62 is greater than the critical value of 60, we would reject the null hypothesis. Or, to phrase it slightly differently, we say that the test has produced a **significant** result. 

### A note on statistical "significance"
>*Like other occult techniques of divination, the statistical method has a private jargon deliberately contrived to obscure its methods from non-practitioners.*
>
>-- Attributed to G. O. Ashley^[The internet seems fairly convinced that Ashley said this, though I can't for the life of me find anyone willing to give a source for the claim.]

A very brief digression is in order at this point, regarding the word "significant". The concept of statistical significance is actually a very simple one, but has a very unfortunate name. If the data allow us to reject the null hypothesis, we say that "the result is *statistically significant*", which is often shortened to "the result is significant". This terminology is rather old, and dates back to a time when "significant" just meant something like "indicated", rather than its modern meaning, which is much closer to "important". As a result, a lot of modern readers get very confused when they start learning statistics, because they think that a "significant result" must be an important one. It doesn't mean that at all. All that "statistically significant" means is that the data allowed us to reject a null hypothesis. Whether or not the result is actually important in the real world is a very different question, and depends on all sorts of other things. 


### The difference between one sided and two sided tests{#onesidedtests}
There's one more thing I want to point out about the hypothesis test that I've just constructed. If we take a moment to think about the statistical hypotheses I've been using, 
$$
\begin{array}{cc}
H_0 : & \theta = .5 \\
H_1 : & \theta \neq .5 
\end{array}
$$
we notice that the alternative hypothesis covers *both* the possibility that $\theta < .5$ and the possibility that $\theta > .5$. This makes sense if I really think that ESP could produce better-than-chance performance *or* worse-than-chance performance (and there are some people who think that). In statistical language, this is an example of a **two-sided test**. It's called this because the alternative hypothesis covers the area on both "sides" of the null hypothesis, and as a consequence the critical region of the test covers both tails of the sampling distribution (2.5\% on either side if $\alpha =.05$), as illustrated earlier in Figure \@ref(fig:crit2). 

However, that's not the only possibility. It might be the case, for example, that I'm only willing to believe in ESP if it produces better than chance performance. If so, then my alternative hypothesis would only covers the possibility that $\theta > .5$, and as a consequence the null hypothesis now becomes $\theta \leq .5$:
$$
\begin{array}{cc}
H_0 : & \theta \leq .5 \\
H_1 : & \theta > .5 
\end{array}
$$
When this happens, we have what's called a **one-sided test**, and when this happens the critical region only covers one tail of the sampling distribution. This is illustrated in Figure \@ref(fig:crit1).

```{r crit1, echo=FALSE, fig.cap="The critical region for a one sided test. In this case, the alternative hypothesis is that $\\theta > .05$, so we would only reject the null hypothesis for large values of $X$. As a consequence, the critical region only covers the upper tail of the sampling distribution; specifically the upper 5% of the distribution. Contrast this to the two-sided version earlier)"}
setUpPlot()
	z <- x>=58
	addDistPlot(x,y,z)
	h <- .03
	addArrow(c(58,80),h)	
	text(75,h+.013,"critical region")
	text(75,h+.007,"(5% of the distribution)")
	title(main="Critical Region for a One-Sided Test",font.main=1)
```

## The $p$ value of a test{#pvalue}
In one sense, our hypothesis test is complete; we've constructed a test statistic, figured out its sampling distribution if the null hypothesis is true, and then constructed the critical region for the test. Nevertheless, I've actually omitted the most important number of all: **the $p$ value**. It is to this topic that we now turn. 

There are actually different ways of interpreting a $p$ value, that reflect very different ways of thinking about hypothesis tests. We will stick to interpreting a $p$ value as is done in most introductory textbooks, but please be aware this is not the only way of thinking about $p$ values. In fact, how the $p$ value is used in practice is a mix of several school of thoughts, which makes the whole thing kind of a mess, but it's a mess we have to deal with.   

### The probability of extreme data
This definition $p$-value comes from Sir Ronald Fisher (yes, [him again](#pears-r)) and is the one that you tend to see in most introductory statistics textbooks. Notice how, when I constructed the critical region, it corresponded to the *tails* (i.e., extreme values) of the sampling distribution? That's not a coincidence: almost all "good" tests have this characteristic (good in the sense of minimising our type II error rate, $\beta$). The reason for that is that a good critical region almost always corresponds to those values of the test statistic that are least likely to be observed if the null hypothesis is true. If this rule is true, then we can define the $p$-value as the probability that we would have observed a test statistic that is at least as extreme as the one we actually did get. In other words, if the data are extremely implausible according to the null hypothesis, then the null hypothesis is probably wrong.

### A common mistake
Unfortunately, there is an explanation of $p$-values that people sometimes give, especially when they're first learning statistics, and it is **absolutely and completely wrong**. This mistaken approach is to refer to the $p$ value as "the probability that the null hypothesis is true". It's an intuitively appealing way to think, but it's wrong in two key respects: (1) null hypothesis testing is a frequentist tool, and the frequentist approach to probability does *not* allow you to assign probabilities to the null hypothesis... according to this view of probability, the null hypothesis is either true or it is not; it cannot have a "5\% chance" of being true. (2) even within the Bayesian approach, which does let you assign probabilities to hypotheses, the $p$ value would not correspond to the probability that the null is true; this interpretation is entirely inconsistent with the mathematics of how the $p$ value is calculated. Put bluntly, despite the intuitive appeal of thinking this way, there is *no* justification for interpreting a $p$ value this way. Never do it.

## Reporting the results of a hypothesis test{#writeup}
When writing up the results of a hypothesis test, there's usually several pieces of information that you need to report, but it varies a fair bit from test to test. Materials with examples for writing up the results of different statistical tests are provided in the basestone project on Canvas. However, regardless of what test you're doing, the one thing that you always have to do is say something about the $p$ value, and whether or not the outcome was significant. 

The fact that you have to do this is unsurprising; it's the whole point of doing the test. What might be surprising is the fact that there is some contention over exactly how you're supposed to do it. Leaving aside those people who completely disagree with the entire framework underpinning null hypothesis testing, there's a certain amount of tension that exists regarding whether or not to report the exact $p$ value that you obtained, or if you should state only that $p < \alpha$ for a significance level that you chose in advance (e.g., $p<.05$). 

The convention is that scientists rely on three standard significance levels: .05, .01 and .001. When reporting your results, you indicate which (if any) of these significance levels allow you to reject the null hypothesis. This is summarised in Table \@ref(tab:pvaltable). 

```{r pvaltable, echo=FALSE}
knitr::kable(data.frame(stringsAsFactors=FALSE,
        Usual.notation = c("$p>.05$", "$p<.05$", "$p<.01$", "$p<.001$"),
         Signif..stars = c("n.s.", "*", "**", "***"),
   English.translation = c("The test wasn't significant",
                           "The test was significant at $\\alpha = .05$ but not at $\\alpha =.01$ or $\\alpha = .001$",
                           "The test was significant at $\\alpha = .05$  and $\\alpha = .01$ but not at  $\\alpha = .001$",
                           "The test was significant at all levels"),
        The.null.is... = c("Retained", "Rejected", "Rejected", "Rejected")
), col.names = c("Usual notation", "Signif. stars", "Plain English", "The null is..."), caption = "A commonly adopted convention for reporting $p$ values: in many places it is conventional to report one of four different things (e.g., $p<.05$) as shown below. I've included the \"significance stars\" notation (i.e., a * indicates $p<.05$) because you sometimes see this notation produced by statistical software.")

```
Quite a lot of people prefer to report exact $p$ values instead. To many people, the advantage of allowing the reader to make up their own mind about how to interpret $p = .06$ outweighs any disadvantages. In practice, however, even among those researchers who prefer exact $p$ values it is quite common to just write $p<.001$ instead of reporting an exact value for small $p$. This is in part because a lot of software doesn't actually print out the $p$ value when it's that small (e.g., SPSS just writes $p = .000$ whenever $p<.001$), and in part because a very small $p$ value can be kind of misleading. The human mind sees a number like .0000000001 and it's hard to suppress the gut feeling that the evidence in favour of the alternative hypothesis is a near certainty. In practice however, this is usually wrong. Life is a big, messy, complicated thing: and every statistical test ever invented relies on simplifications, approximations and assumptions. As a consequence, it's probably not reasonable to walk away from *any* statistical analysis with a feeling of confidence stronger than $p<.001$ implies. In other words, $p<.001$ is really code for "as far as *this test* is concerned, the evidence is overwhelming." 

In light of all this, you might be wondering exactly what you should do. There's a fair bit of contradictory advice on the topic, with some people arguing that you should report the exact $p$ value, and other people arguing that you should use the tiered approach illustrated in Table \@ref(tab:pvaltable). As a result, the best advice I can give is to suggest that you look at papers/reports written in your field and see what the convention seems to be. If there doesn't seem to be any consistent pattern, then use whichever method you prefer. 

## Running the hypothesis test in practice
At this point some of you might be wondering if this is a "real" hypothesis test, or just a toy example that I made up. It's real. In the previous discussion I built the test from first principles, thinking that it was the simplest possible problem that you might ever encounter in real life. However, this test already exists: it's called the **binomial test**, and it's implemented by an R function called `binom.test()`. To test the null hypothesis that the response probability is 0.5 (`p = .5`),^[Note that the `p` here has nothing to do with a $p$ value. The `p` argument in the `binom.test()` function corresponds to the probability of making a correct response, according to the null hypothesis. In other words, it's the $\theta$ value.] using data in which `x = 62` of `n = 100` people made the correct response, here's how to do it in R:

```{r,echo=T}
binom.test( x=62, n=100, p=.5 )
```
Right now, this output looks pretty unfamiliar to you, but you can see that it's telling you more or less the right things. Specifically, the $p$-value of 0.02 is less than the usual choice of $\alpha = .05$, so you can reject the null. We'll talk a lot more about how to read this sort of output as we go along; and after a while you'll hopefully find it quite easy to read and understand. For now, however, I just wanted to make the point that R contains a whole lot of functions corresponding to different kinds of hypothesis test. And while I'll usually spend quite a lot of time explaining the logic behind how the tests are built, every time I discuss a hypothesis test the discussion will end with me showing you a fairly simple R command that you can use to run the test in practice.

## Some issues to consider{#nhstmess}
What I've described to you in this chapter is the orthodox framework for null hypothesis significance testing (NHST). Understanding how NHST works is an absolute necessity, since it has been the dominant approach to inferential statistics ever since it came to prominence in the early 20th century. It's what the vast majority of working scientists rely on for their data analysis, so even if you hate it you need to know it. However, the approach is not without problems. There are a number of quirks in the framework, historical oddities in how it came to be, theoretical disputes over whether or not the framework is right, and a lot of practical traps for the unwary. I'm not going to go into a lot of detail on this topic, but I think it's worth briefly discussing a few of these issues.

### Neyman versus Fisher
The first thing you should be aware of is that orthodox NHST is actually a mash-up of two rather different approaches to hypothesis testing, one proposed by Sir Ronald Fisher and the other proposed by Jerzy Neyman [for a historical summary see @lehmann2011fisher]. The history is messy because Fisher and Neyman were real people whose opinions changed over time, and at no point did either of them offer "the definitive statement" of how we should interpret their work many decades later. That said, here's a quick summary of what I take these two approaches to be. 

First, let's talk about Fisher's approach. As far as I can tell, Fisher assumed that you only had the one hypothesis (the null), and what you want to do is find out if the null hypothesis is inconsistent with the data. From his perspective, what you should do is check to see if the data are "sufficiently unlikely" according to the null. In fact, if you remember back to our earlier discussion, that's how Fisher defines the $p$-value. According to Fisher, if the null hypothesis provided a very poor account of the data, you could safely reject it. But, since you don't have any other hypotheses to compare it to, there's no way of "accepting the alternative" because you don't necessarily have an explicitly stated alternative. That's more or less all that there was to it. 

In contrast, Neyman thought that the point of hypothesis testing was as a guide to action, and his approach was somewhat more formal than Fisher's. His view was that there are multiple things that you could *do* (accept the null or accept the alternative) and the point of the test was to tell you which one the data support. From this perspective, it is critical to specify your alternative hypothesis properly. If you don't know what the alternative hypothesis is, then you don't know how powerful the test is, or even which action makes sense. His framework genuinely requires a competition between different hypotheses. For Neyman, the $p$ value didn't directly measure the probability of the data (or data more extreme) under the null, it was more of an abstract description about which "possible tests" were telling you to accept the null, and which "possible tests" were telling you to accept the alternative.

As you can see, what we have today is an odd mishmash of the two. We talk about having both a null hypothesis and an alternative (Neyman), but usually define the $p$ value in terms of extreme data (Fisher), but we still have $\alpha$ values (Neyman). Some of the statistical tests have explicitly specified alternatives (Neyman) but others are quite vague about it (Fisher). And, according to some people at least, we're not allowed to talk about accepting the alternative (Fisher). It's a mess: but I hope this at least explains why it's a mess.

### Bayesians versus frequentists
Earlier on in this chapter I was quite emphatic about the fact that you *cannot* interpret the $p$ value as the probability that the null hypothesis is true. NHST is fundamentally a frequentist tool (see Chapter \@ref(chance)) and as such it does not allow you to assign probabilities to hypotheses: the null hypothesis is either true or it is not. The Bayesian approach to statistics interprets probability as a degree of belief, so it's totally okay to say that there is a 10\% chance that the null hypothesis is true: that's just a reflection of the degree of confidence that you have in this hypothesis. You aren't allowed to do this within the frequentist approach. Remember, if you're a frequentist, a probability can only be defined in terms of what happens after a large number of independent replications (i.e., a long run frequency). If this is your interpretation of probability, talking about the "probability" that the null hypothesis is true is complete gibberish: a null hypothesis is either true or it is false. There's no way you can talk about a long run frequency for this statement. To talk about "the probability of the null hypothesis" is as meaningless as "the color of freedom". It doesn't have one!

Most importantly, this *isn't* a purely ideological matter. If you decide that you are a Bayesian and that you're okay with making probability statements about hypotheses, you have to follow the Bayesian rules for calculating those probabilities. You can find more on [Bayesian statistics in other sources](https://sites.google.com/site/doingbayesiandataanalysis/), but for now what I want to point out to you is the $p$ value is a *terrible* approximation to the probability that $H_0$ is true. If what you want to know is the probability of the null, then the $p$ value is not what you're looking for!

### Traps
As you can see, the theory behind hypothesis testing is a mess, and even now there are arguments in statistics about how it "should" work. However, disagreements among statisticians are not our real concern here. Our real concern is practical data analysis. And while the "orthodox" approach to null hypothesis significance testing has many drawbacks, even an unrepentant Bayesian like myself would agree that they can be useful if used responsibly. Most of the time they give sensible answers, and you can use them to learn interesting things. Setting aside the various ideologies and historical confusions that we've discussed, the fact remains that the biggest danger in all of statistics is *thoughtlessness*. I don't mean stupidity, here: I literally mean thoughtlessness. The rush to interpret a result without spending time thinking through what each test actually says about the data, and checking whether that's consistent with how you've interpreted it. That's where the biggest trap lies. 
To give an example of this, consider the following example [see @gelman2006difference]. Suppose I'm running my ESP study, and I've decided to analyse the data separately for the male participants and the female participants. Of the male participants, 33 out of 50 guessed the colour of the card correctly. This is a significant effect ($p = .03$). Of the female participants, 29 out of 50 guessed correctly. This is not a significant effect ($p = .32$). Upon observing this, it is extremely tempting for people to start wondering why there is a difference between males and females in terms of their psychic abilities. However, this is wrong. If you think about it, we haven't *actually* run a test that explicitly compares males to females. All we have done is compare males to chance (binomial test was significant) and compared females to chance (binomial test was non significant). If we want to argue that there is a real difference between the males and the females, we should probably run a test of the null hypothesis that there is no difference! We can do that using a different hypothesis test,^[In this case a Pearson chi-square test of independence] but when we do that it turns out that we have no evidence that males and females are significantly different ($p = .54$). *Now* do you think that there's anything fundamentally different between the two groups? Of course not. What's happened here is that the data from both groups (male and female) are pretty borderline: by pure chance, one of them happened to end up on the magic side of the $p = .05$ line, and the other one didn't. That doesn't actually imply that males and females are different. This mistake is so common that you should always be wary of it: the difference between significant and not-significant is *not* evidence of a real difference -- if you want to say that there's a difference between two groups, then you have to test for that difference! 

The example above is just that: an example. I've singled it out because it's such a common one, but the bigger picture is that data analysis can be tricky to get right. Think about *what* it is you want to test, *why* you want to test it, and whether or not the answers that your test gives could possibly make any sense in the real world. 

## That's it for this week!
Null hypothesis significance testing (NHST) is one of the most ubiquitous elements to statistical theory. The vast majority of scientific papers report the results of some hypothesis test or another. As a consequence it is almost impossible to get by in science without having at least a cursory understanding of NHST and what a $p$-value means, but we also learned there are some issues with how these concepts are used in practice. 

During the labs we will walk through the formal steps of conducting a NHST, use R to run tests on our own data and discuss some of things to be mindful of when conducting a statistical test. 